---
title: "BST 260 Group Project"
author: "Christina, Dena, Jack, Veronica, Whitney"
date: "11/22/2021"
output:
  html_document: default
  pdf_document: default
---

```{r}
#some basic packages that we will probably want: 
library(tidyverse)
library(lubridate)
library(rvest)

#package to read in xlsx files from a url:
library(openxlsx)
```

# Analysis 1: Data scraping, cleaning and merging

Obtain County Health Ranking Data sheets, as well as county-level COVID data from NYT:

```{r}
#County Health Ranking Data:
dat <- read.xlsx("https://www.countyhealthrankings.org/sites/default/files/media/document/2021%20County%20Health%20Rankings%20Data%20-%20v1.xlsx", sheet = 4, startRow = 2)

#County Health Ranking Data second sheet:
dat2 <- read.xlsx("https://www.countyhealthrankings.org/sites/default/files/media/document/2021%20County%20Health%20Rankings%20Data%20-%20v1.xlsx", sheet = 6, startRow = 2)

#Cumulative County-level COVID data:
covid_dat <- read.csv(url("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"))
```

Web scraping for County names and population from Wikipedia:

```{r}

url <- "https://en.wikipedia.org/wiki/List_of_United_States_counties_and_county_equivalents"
tab = read_html(url) %>% html_nodes("table")

county_names <- c("county", "state", "population_2019", "delete")
counties <- tab %>% .[1] %>% html_table %>% .[[1]] %>% setNames(county_names)

counties <- counties %>% 
  select(!delete)

#make population_2019 numeric:
counties$population_2019 <- as.numeric(gsub(",","", counties$population_2019))

```

Data Manipulation and Merging:

```{r}
#Filter COVID data to most recent date, then arrange by state:
covid_dat <- covid_dat %>%
  filter(date == '2021-11-21') %>%
  select(!fips) %>%
  arrange(state)

#Make a list of the column names that we're interested in for dat:
columns <- c("FIPS",	"State",	"County", "Years.of.Potential.Life.Lost.Rate", "%.Fair.or.Poor.Health", "Average.Number.of.Physically.Unhealthy.Days",	"Average.Number.of.Mentally.Unhealthy.Days",	"%.Low.birthweight",	"%.Smokers", "%.Adults.with.Obesity",	"Food.Environment.Index",	"%.Physically.Inactive",	"%.With.Access.to.Exercise.Opportunities",	"%.Excessive.Drinking",	"#.Alcohol-Impaired.Driving.Deaths",	"#.Driving.Deaths",	"%.Driving.Deaths.with.Alcohol.Involvement",	"Chlamydia.Rate",	"Teen.Birth.Rate",	"%.Uninsured", "Primary.Care.Physicians.Rate",	"Primary.Care.Physicians.Ratio",	"Dentist.Rate",	"Dentist.Ratio",	"Mental.Health.Provider.Rate",	"Mental.Health.Provider.Ratio",	"Preventable.Hospitalization.Rate",	"%.With.Annual.Mammogram",	"%.Vaccinated",	"%.Completed.High.School",	"%.Some.College",	"%.Unemployed",	"%.Children.in.Poverty",	"Income.Ratio",	"%.Children.in.Single-Parent.Households",	"Social.Association.Rate", "Violent.Crime.Rate",	"Injury.Death.Rate",	"Average.Daily.PM2.5",	"Presence.of.Water.Violation", "%.Severe.Housing.Problems",	"%.Drive.Alone.to.Work", "%.Long.Commute.-.Drives.Alone")

#Keep only the columns from our list above, rename to match COVID data, arrange by state:
dat <- dat %>%
  select(columns) %>%
  rename(county = County) %>%
  rename(state = State) %>%
  rename(fips = FIPS) %>%
  arrange(state)

#Make a list of the column names we're interested in for dat2:
columns2 <- c("FIPS",	"State",	"County", "Life.Expectancy",	"Age-Adjusted.Death.Rate",	"Child.Mortality.Rate",	"Infant.Mortality.Rate",	"%.Frequent.Physical.Distress",	"%.Frequent.Mental.Distress",	"%.Adults.with.Diabetes",	"HIV.Prevalence.Rate",	"%.Food.Insecure",	"%.Limited.Access.to.Healthy.Foods",	"Drug.Overdose.Mortality.Rate",	"Motor.Vehicle.Mortality.Rate",	"%.Insufficient.Sleep",	"%.Disconnected.Youth",	"Average.Grade.Performance",	"Median.Household.Income",	"%.Enrolled.in.Free.or.Reduced.Lunch",	"Homicide.Rate",	"Suicide.Rate.(Age-Adjusted)",	"Firearm.Fatalities.Rate",	"Juvenile.Arrest.Rate",	"%.Homeowners",	"%.Broadband.Access",	"%.Female",	"%.Rural")

#Keep only the columns from our list above, rename to match COVID data, arrange by state:
dat2 <- dat2 %>%
  select(columns2) %>%
  rename(county = County) %>%
  rename(state = State) %>%
  rename(fips = FIPS) %>%
  arrange(state)

#Merge dat with dat2, then merge with COVID data:
dat <- dat %>%
  inner_join(dat2, by=c("county", "state"))

dat <- dat %>%
  left_join(covid_dat, by=c("county", "state"))

#Finally, merge dat with counties for population:
dat <- dat %>%
  left_join(counties, by=c("county", "state"))

```

# Analysis 2: Data Visualizations 

## Analysis 2.2: Boxplot of median household income per state

```{r}
##box plot of household income per state 
p <- dat %>% 
  mutate(state = reorder(state, Median.Household.Income, FUN = median)) %>%
  ggplot(aes(state, Median.Household.Income, fill = state)) +
  geom_boxplot() + 
  theme_bw() +
  ggtitle("Boxplots of Median Household Income by State")  +
  theme_update(plot.title = element_text(hjust = 0.6)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("State") + ylab("Median Household Income")+
  theme(legend.position = "none")
p 


```

## Analysis 2.3: Life Expectancy

```{r}
#box plot of life expectancy by state
p2 <- dat %>% 
  mutate(state = reorder(state, Life.Expectancy, FUN = median), is.na(Life.Expectancy)) %>%
  ggplot(aes(x = reorder(state, Life.Expectancy), y = Life.Expectancy, fill = state)) +
  geom_boxplot() + 
  theme_bw() +
  theme(legend.position = "none") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  ggtitle("Boxplots of Life Expectancy by State") + 
  xlab("State") + ylab("Life Expectancy (years)")

p2


###Extra code to assess individual state statistics
#something <- dat %>% select(state, Life.Expectancy, county) %>% group_by(state) %>%
#                mutate(median_life = median(Life.Expectancy)) %>% ungroup()
#something

#something2 <- dat %>% filter(state == "Mississippi", !is.na(Life.Expectancy))
##something2
#median(something2$Life.Expectancy)

#code to determine distribution of life expectancy variable 
#mean(dat$Life.Expectancy, na.rm = TRUE)
#min(dat$Life.Expectancy, na.rm = TRUE)
#summary(dat$Life.Expectancy)


#selecting the top 24 counties with the lowest life expectancy
p3 <- dat %>% select(county, state, Life.Expectancy) %>% arrange(Life.Expectancy)%>% 
          filter(Life.Expectancy <= 69.7)

#creating a line plot of descending order of life expectancy
p4 <- p3 %>% ggplot(aes(x = reorder(county, -Life.Expectancy), y = Life.Expectancy, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Shortest Life Expectancy in the United States") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Life Expectancy (years)") 

#adding state abbreviations to county line plot
p4 + scale_x_discrete("U.S. County", labels = c("Oglala Lakota" = "Oglala Lakota (SD)", "Buffalo" = " Buffalo (SD)", "Corson" = "Corson (SD)", "Mellette" = " Mellette (SD)", "Covington City" = "Covington City (VA)" , "Kusilvak" = "Kusilvak (AK)", "Todd" = " Todd (SD)", "Petersburg City" = "Petersburg City (VA)", "Galax City" = "Galax City (VA)", "Roosevelt" = "Roosevelt (MT)", "Martinsville City" = "Martinsville City (VA)", "Dewey" = "Dewey (SD)", "Big Horn" = "Big Horn (MT)", "Union" = "Union (FL)", "Sioux" = "Sioux (ND)", "Perry" = "Perry (KY)",  "Kingman" = "Kingman (KS)", "McDowell" = "McDowell (WV)", "Edwards" = "Edwards (KS)", "Bennett" = "Bennett (SD)", "Owsley" = "Owsley (KY)", "Thurston" = "Thurston (NE)", "Breathitt" = "Breathitt (KY)", "Okfuskee" = "Okfuskee (OK)", "Logan" = "Logan (WV)"))

##histogram of life expectancy in the united states
p5 <- dat %>% ggplot(aes(x = Life.Expectancy)) +
              geom_histogram(binwidth = 0.2, color = "black", fill = "light blue") + 
              theme_bw()+ 
              ggtitle("Life Expectancy Distribution in the United States") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Life Expectancy (years)") 

p5 + geom_vline(aes(xintercept = mean(Life.Expectancy, na.rm = TRUE)), col='dark red',size=1)

#Code for distribution statistics 
#sd(dat$Life.Expectancy, na.rm = TRUE)
#mean(dat$Life.Expectancy, na.rm = TRUE)


```

## Analysis 2.4: Drug overdose mortality rate 

```{r}
#####Drug Overdose Mortality
#Histogram for drug overdose Mortality variable 
p6 <- dat %>% ggplot(aes(x = `Drug.Overdose.Mortality.Rate`)) +
              geom_histogram(binwidth = 0.75, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Drug Overdose Mortality Rates per 100,000 population in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Drug Overdose Mortality Rate (per 100,000 population)") 
p6 + geom_vline(aes(xintercept = median(`Drug.Overdose.Mortality.Rate`, na.rm = TRUE)), col='dark red',size=1) 

#Code for distribution statistics 
#median(dat$`Drug.Overdose.Mortality.Rate`, na.rm = TRUE)
#IQR(dat$`Drug.Overdose.Mortality.Rate`, na.rm = TRUE)
#max(dat$`Drug.Overdose.Mortality.Rate`, na.rm = TRUE)

#boxplot of drug overdose mortality rate by state
p7 <- dat %>% 
  ggplot(aes(x = reorder(state, Drug.Overdose.Mortality.Rate) , y = Drug.Overdose.Mortality.Rate, fill = state))+
  geom_boxplot() + 
  theme_bw() +
  ggtitle("Boxplots of Drug Overdose Mortality Rate by State")  +
  theme_update(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5)) +
  theme(axis.text.y = element_text(size = 0.1)) + ##need to adjust the size of text 
  xlab("State") + ylab("Drug Overdose Mortality Rate per 100,000 Pop.")+ #change size of text
  theme(legend.position = "none")

p7 
###Extra code to assess individual state statistics
#something2 <- dat %>% filter(state == "West Virginia", !is.na(Drug.Overdose.Mortality.Rate))
#something2

#code to determine the top 25 counties with highest drug overdose mortality rate
p8 <- dat %>% select(county, state, Drug.Overdose.Mortality.Rate) %>% arrange(Drug.Overdose.Mortality.Rate) %>%
          filter(Drug.Overdose.Mortality.Rate >= 61) 

#creating a line plot of ascending order of drug overdose mortality rate
p9 <- p8 %>% ggplot(aes(x = reorder(county, Drug.Overdose.Mortality.Rate), y = Drug.Overdose.Mortality.Rate, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            theme(axis.text.y = element_text(size = 0.5)) +
            ggtitle("Top 25 Counties With the Highest Drug Overdose Mortality Ratio in the U.S.") +
            xlab("U.S County") + ylab("Drug Overdose Mortality Ratio (per 100,000 pop.)") 

#adding state abbreviations to county line plot
p9 + scale_x_discrete("U.S. County", labels = c("Kenton" = "Kenton (KY)", "Lake" = " Lake (CA)", "Cheatham" = "Cheatham (TN)", "Hampshire" = " Hampshire (WV)", "Owen" = "Owen (KY)", "Gallia" = "Gallia (OH)", "Lincoln" = "Lincoln (WV)", "Boyd" = "Boyd (KY)", "Union" = "Union (IN)", "Estill" = "Estill (KY)", "Fayette" = "Fayette (WV) & (IN)", "Mahnomen" = "Mahnomen (MN)", "Cecil" = "Cecil (MD)", "Montgomery" = "Montgomery (OH)", "Kanawha" = "Kanawha (WV)", "Gallatin" = "Gallatin (KY)", "Boone" = "Boone (WV)", "Rio Arriba" = "Rio Arriba (NM)", "Raleigh" = "Raleigh (WV)", "Wayne" = "Wayne	(IN)", "Scioto" = "Scioto (OH)", "Logan" = "Logan (WV)", "Wayne" = "Wayne (WV)", "Cabell" = "Cabell (WV)", "Berkeley" = "Berkeley (WV)"))


```

##Analysis 2.5: Scatter plot of life expectancy vs. drug over dose mortality rate

```{r}
###Scatter plot of drug over dose mortality and life expectancy 
p10 <- dat %>% ggplot(aes(x = Drug.Overdose.Mortality.Rate, y = Life.Expectancy)) +
            geom_point(size = 1) + 
            ggtitle("Scatter Plot of Life Expectancy vs. Drug Overdose Mortality Rate (per 100,000 population)") +
            theme(plot.title = element_text(size = 10)) + 
            xlab("Drug Overdose Mortality Rate (per 100,000 population)") + 
            ylab("Life Expectancy (years)") 

#adding a trend line to the scatter plot     
p10 + geom_smooth(method = "lm")

```

## Analysis 2.6: Food Insecurity

```{r}
####Food Insecurity
#histogram of food insecurity
p_food <- dat %>% ggplot(aes(x = `%.Food.Insecure`)) +
              geom_histogram(binwidth = 0.75, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Food Insecurity % in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Food Insecurity %") 

#code to add line where the mean value is located 
p_food + geom_vline(aes(xintercept = mean(`%.Food.Insecure`, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#mean(dat$`%.Food.Insecure`, na.rm = TRUE)
#sd(dat$`%.Food.Insecure`, na.rm = TRUE)

#selecting the top 25 counties with the highest food insecurity %
p11 <- dat %>% select(county, state, `%.Food.Insecure`) %>% arrange(`%.Food.Insecure`)%>% 
          filter(`%.Food.Insecure` >= 23.9)

#creating a line plot of ascending order of food insecurity %
p12 <- p11 %>% ggplot(aes(x = reorder(county, `%.Food.Insecure`), y = `%.Food.Insecure`, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Highest Food Insecurity % in the United States") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Food Insecurity %") 

#adding state abbreviations to county line plot
p12 + scale_x_discrete("U.S. County", labels = c("Chicot" = "Chicot (AR)", "Coahoma" = "Coahoma (MS)", "Bolivar" = "Bolivar (MS)", "Clay" = " Clay (GA) & Clay (KY)", "Leflore" = "Leflore (MS)", "Wolfe" = "Wolfe (KY)", "Washington" = "Washington (MS)", "Breathitt" = "Breathitt (KY)", "Wilcox" = "Wilcox (AL)", "Bell" = "Bell (KY)", "Magoffin" = "Magoffin (KY)", "Harlan" = "Harlan (KY)", "Wilkinson" = "Wilkinson (MS)", "Quitman" = "Quitman (MS)", "Phillips" = "Phillips (AR)", "Todd" = "Todd (SD)", "Greene" = "Greene (AL)", "Humphreys" = "Humphreys (MS)", "Oglala Lakota" = "Oglala Lakota (SD)", "Perry" = "Perry (AL)", "Claiborne" = "Claiborne (MS)", "Holmes" = "Holmes (MS)", "Issaquena" = "Issaquena (MS)", "Jefferson" = "Jefferson (MS)"))

```

## Analysis 2.7: Severe Housing Problems 

```{r}
#histogram of severe housing problems
p_housing <- dat %>% ggplot(aes(x = `%.Severe.Housing.Problems`)) +
              geom_histogram(binwidth = 0.75, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Severe Housing % in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Severe Housing Probelms %") 

#code to add line where the mean value is located 
p_housing + geom_vline(aes(xintercept = mean(`%.Severe.Housing.Problems`, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#mean(dat$`%.Severe.Housing.Problems`, na.rm = TRUE)
#sd(dat$`%.Severe.Housing.Problems`, na.rm = TRUE)

#selecting the top 25 counties with the highest severe housing problems %
p13 <- dat %>% select(county, state, `%.Severe.Housing.Problems`) %>% arrange(`%.Severe.Housing.Problems`)%>% 
          filter(`%.Severe.Housing.Problems` >= 26.43805)

#creating a line plot of ascending order of housing problems in top 25 counties 
p14 <- p13 %>% ggplot(aes(x = reorder(county, `%.Severe.Housing.Problems`), y = `%.Severe.Housing.Problems`, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Highest Severe Housing Problems % in the United States") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Severe Housing Problems %") 

#adding state abbreviations to county line plot
p14 + scale_x_discrete("U.S. County", labels = c("Imperial" = "Imperial (CA)", "Mendocino" = "Mendocino (CA)", "Tunica" = "Tunica (MS)", "Rockland" = " Rockland (NY)", "Orange" = "Orange (CA)", "Ziebach" = "Ziebach (SD)", "Fresno" = "Fresno (CA)", "Tulare" = "Tulare (CA)", "Santa Cruz" = "Santa Cruz (CA)", "Todd" = "Todd (SD)", "Essex" = "Essex (NJ)", "Alpine" = "Alpine (CA)", "Hudson" = "Hudson (NJ)", "San Miguel" = "San Miguel (CO)", "McKinley" = "McKinley (NM)", "Webb" = "Webb (TX)", "Santa Barbara" = "Santa Barbara (CA)", "Buffalo" = "Buffalo (SD)", "Monterey" = "Monterey (CA)", "Miami-Dade" = "Miami-Dade (FL)", "Passaic" = "Passaic (NJ)", "Apache" = "Apache (AZ)", "Madison" = "Madison (ID)", "Los Angeles" = "Los Angeles (CA)", "Oglala Lakota" = "Oglala Lakota	(SD)"))

```

## Analysis 2.8: Some College 

```{r}
#histogram of some college variable 
p_college <- dat %>% ggplot(aes(x = `%.Some.College`)) +
              geom_histogram(binwidth = 0.75, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Some College % in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Some College %") 

#code to add line where the mean value is located 
p_college + geom_vline(aes(xintercept = mean(`%.Some.College`, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#mean(dat$`%.Some.College`, na.rm = TRUE)
#sd(dat$`%.Some.College`, na.rm = TRUE)


#selecting the top 25 counties with the lowest some college %
p14 <- dat %>% select(county, state, `%.Some.College`) %>% arrange(`%.Some.College`)%>%
          filter(`%.Some.College` < 29.86)

#line plot for uninsured in the top 25 counties
p15 <- p14 %>% ggplot(aes(x = reorder(county, -`%.Some.College`), y = `%.Some.College`, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Lowest Some College % in the United States") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Some College %") 

#adding state abbreviations to county line plot
p15 + scale_x_discrete("U.S. County", labels = c("Kenedy" = "Kenedy (TX)", "Jeff Davis" = "Jeff Davis (TX)", "La Salle" = "La Salle (TX)", "Forest" = " Forest (PA)", "Holmes" = "Holmes (OH)", "Taliaferro" = "Taliaferro (GA)", "Kusilvak" = "Kusilvak (AK)", "DeSoto" = "DeSoto (FL)", "Hudspeth" = "Hudspeth (TX)", "Lincoln" = "Lincoln (CO)", "Hancock" = "Hancock (GA)", "Tensas" = "Tensas (LA)", "McDowell" = "McDowell (WV)", "Garza" = "Garza (TX)", "Telfair" = "Telfair (GA)", "Hamilton" = "Hamilton (FL)", "Hardee" = "Hardee (FL)", "Stewart" = "Stewart (GA)", "East Carroll" = "East Carroll (LA)", "LaGrange" = "LaGrange (IN)", "Bent" = "Bent (CO)", "Real" = "Real (TX)", "Glades" = "Glades (FL)", "Issaquena" = "Issaquena (MS)", "Wolfe" = "Wolfe (KY)", "Gaines" = "Gaines (TX)", "Lake" = "Lake (TN)", "Menard" = "Menard	(TX)"))

```

## Analysis 2.9: Uninsured

```{r}
#histogram of uninsured variable 
p_uninsured <- dat %>% ggplot(aes(x = `%.Uninsured`)) +
              geom_histogram(binwidth = 0.75, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Uninsured % in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Uninsured %") 

#code to add line where the median value is located 
p_uninsured + geom_vline(aes(xintercept = median(`%.Uninsured`, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#median(dat$`%.Uninsured`, na.rm = TRUE)
#IQR(dat$`%.Uninsured`, na.rm = TRUE)

#selecting the top 25 counties with the highest uninsured %
p16 <- dat %>% select(county, state, `%.Uninsured`) %>% arrange(desc(`%.Uninsured`))%>%
          filter(`%.Uninsured` > 27.1)

#line plot for uninsured in the top 25 counties
p17 <- p16 %>% ggplot(aes(x = reorder(county, `%.Uninsured`), y = `%.Uninsured`, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Highest Uninsured % in the United States") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Uninsured %") 

#adding state abbreviations to county line plot
p17 + scale_x_discrete("U.S. County", labels = c("Gaines" = "Gaines (TX)", "Hidalgo" = "Hidalgo (TX)", "Presidio" = "Presidio (TX)", "Cameron" = " Cameron (TX)", "Dallam" = "Dallam (TX)", "Collingsworth" = "Collingsworth (TX)", "Starr" = "Starr (TX)", "Webb" = "Webb (TX)", "Jeff Davis" = "Jeff Davis (TX)", "Bailey" = "Bailey (TX)", "Castro" = "Castro (TX)", "Zapata" = "Zapata (TX)", "Hall" = "Hall (TX)", "Menard" = "Menard (TX)", "Sherman" = "Sherman (TX)", "Hudspeth" = "Hudspeth (TX)", "Parmer" = "Parmer (TX)", "San Saba" = "San Saba (TX)", "Lipscomb" = "Lipscomb (TX)", "Hansford" = "Hansford (TX)", "Maverick" = "Maverick (TX)", "Mason" = "Mason (TX)", "Hendry" = "Hendry (FL)", "Ochiltree" = "Ochiltree (TX)", "Clark" = "Clark	(ID)"))
```

## Analysis 2.10: Unemployment 

```{r}
#histogram of unemployment variable 
p_unemployed <- dat %>% ggplot(aes(x = `%.Unemployed`)) +
              geom_histogram(binwidth = 0.75, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Unemployed % in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Unemployed %") 

#code to add line where the median value is located 
p_unemployed + geom_vline(aes(xintercept = median(`%.Unemployed`, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#median(dat$`%.Unemployed`, na.rm = TRUE)
#IQR(dat$`%.Unemployed`, na.rm = TRUE)

#selecting the top 25 counties with the highest violence crime rate 
p18 <- dat %>% select(county, state, `%.Unemployed`) %>% arrange(desc(`%.Unemployed`))%>%
              filter(`%.Unemployed` > 8.6)

#line plot for violent crime rate in the top 25 counties
p19 <- p18 %>% ggplot(aes(x = reorder(county, `%.Unemployed`), y = `%.Unemployed`, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Highest Unemployed % in the United States") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Unemployment %") 

#adding state abbreviations to county line plot
p19 + scale_x_discrete("U.S. County", labels = c("Imperial" = "Imperial (CA)", "Yuma" = "Yuma (AZ)", "Jefferson" = "Jefferson (MS)", "Calhoun" = "Calhoun (WV)", "Colusa" = "Colusa (CA)", "Luna" = "Luna (NM)", "Ferry" = "Ferry (WA)", "Issaquena" = "Issaquena (MS)", "Magoffin" = "Magoffin (KY)", "Claiborne" = "Claiborne (MS)", "Holmes" = "Holmes (MS)", "Wilkinson" = "Wilkinson (MS)", "Humphreys" = "Humphreys (MS)", "Apache" = "Apache (AZ)", "Starr" = "Starr (TX)", "Oglala Lakota" = "Oglala Lakota (SD)", "Tulare" = "Tulare (CA)", "Harlan" = "Harlan (KY)", "Roane" = "Roane (WV)", "Telfair" = "Telfair (GA)", "Mackinac" = "Mackinac (MI)", "McDowell" = "McDowell (WV)", "Lewis" = "Lewis (KY)", "Sunflower" = "Sunflower (MS)", "Cheboygan" = "Cheboygan	(MI)"))
```

## Analysis 2.11: Violent Crime Rate 

```{r}
#histogram of violent crime rate variable 
p_violence <- dat %>% ggplot(aes(x = `Violent.Crime.Rate`)) +
              geom_histogram(binwidth = 75, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Violent Crime Rate in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Violent Crime Rate") 
#code to add line where the mean value is located 
p_violence + geom_vline(aes(xintercept = median(`Violent.Crime.Rate`, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#median(dat$`Violent.Crime.Rate`, na.rm = TRUE)
#IQR(dat$`Violent.Crime.Rate`, na.rm = TRUE)

#selecting the top 25 counties with the highest violence crime rate 
p20 <- dat %>% select(county, state, `Violent.Crime.Rate`) %>% arrange(desc(`Violent.Crime.Rate`))%>%
              filter(`Violent.Crime.Rate` > 894)

#line plot for violent crime rate in the top 25 counties
p21 <- p20 %>% ggplot(aes(x = reorder(county, `Violent.Crime.Rate`), y = `Violent.Crime.Rate`, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Highest Violent Crime Rate in the United States") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Rate per 100,000 population") 

#adding state abbreviations to county line plot
p21 + scale_x_discrete("U.S. County", labels = c("Calhoun" = "Calhoun (GA)", "Shelby" = "Shelby (TN)", "Alexander" = "Alexander (IL)", "Turner" = "Turner (GA)", "Crittenden" = "Crittenden (AR)", "Marion" = "Marion (IN)", "Phillips" = "Phillips (AR)", "Davidson" = "Davidson (TN)", "Haywood" = "Haywood (TN)", "Levy" = "Levy (FL)", "Pulaski" = "Pulaski (AR)", "Milwaukee" = "Milwaukee (WI)", "Wayne" = "Wayne (MI)", "Jefferson" = "Jefferson (AR)", "Mississippi" = "Mississippi (AR)", "Kenedy" = "Kenedy (TX)", "Bernalillo" = "Bernalillo (NM)", "Jackson" = "Jackson (MO)", "Dougherty" = "Dougherty (GA)", "Scott" = "Scott (MO)", "Dallas" = "Dallas (AL)", "Valencia" = "Valencia (NM)", "Taylor" = "Taylor (FL)", "Dillon" = "Dillon (SC)", "Marlboro" = "Marlboro (SC)"))


```

## Analysis 2.12: Insufficient Sleep

```{r}
#histogram of insufficient sleep variable 
p_sleep <- dat %>% ggplot(aes(x = `%.Insufficient.Sleep`)) +
              geom_histogram(binwidth = 1, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Insufficient Sleep % in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Insufficient Sleep %") 

#code to add line where the mean value is located 
p_sleep + geom_vline(aes(xintercept = mean(`%.Insufficient.Sleep`, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#mean(dat$`%.Insufficient.Sleep`, na.rm = TRUE)
#sd(dat$`%.Insufficient.Sleep`, na.rm = TRUE)

#selecting the top 25 counties with the highest values of insufficient sleep 
p22 <- dat %>% select(county, state, `%.Insufficient.Sleep`) %>% arrange(desc(`%.Insufficient.Sleep`))%>%
              filter(`%.Insufficient.Sleep` > 45.7)

#line plot for insufficient sleep in the top 25 counties 
p23 <- p22 %>% ggplot(aes(x = reorder(county, `%.Insufficient.Sleep`), y = `%.Insufficient.Sleep`, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Highest Insufficient Sleep % in the United States") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Insufficient Sleep %") 

#adding state abbreviations to county line plot
p23 + scale_x_discrete("U.S. County", labels = c("McCreary" = "McCreary (KY)", "Mingo" = "Mingo (WV)", "Greene" = "Greene (AL)", "Boone" = "Boone (WV)", "Pike" = "Pike (KY)", "Hardeman" = "Hardeman (TN)", "McDowell" = "McDowell (WV)", "Clay" = "Clay (KY)", "Letcher" = "Letcher (KY)", "Harlan" = "Harlan (KY)", "Wayne" = "Wayne (WV)", "Clay" = "Clay (GA) and (KY)", "Lowndes" = "Lowndes (AL)", "Logan" = "Logan (WV)", "Lauderdale" = "Lauderdale (TN)", "Summers" = "Summers (WV)", "Jefferson" = "Jefferson (MS)", "Haywood" = "Haywood (TN)", "Martin" = "Martin (KY)", "Hancock" = "Hancock (GA)", "Shelby" = "Shelby (TN)", "Lake" = "Lake (TN)", "Macon" = "Macon (GA)", "Boyd" = "Boyd (KY)", "Breathitt" = "Breathitt (KY)"))
```

## Analysis 2.13: Average Daily PM 2.5 

```{r}
#histogram of average daily PM 2.5 values 
p_pm <- dat %>% ggplot(aes(x = Average.Daily.PM2.5)) +
              geom_histogram(binwidth = 1, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Average Daily PM 2.5 in the U.S.") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Average Daily 2.5") 

#code to add line where the median value is located 
p_pm + geom_vline(aes(xintercept = median(Average.Daily.PM2.5, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#median(dat$Average.Daily.PM2.5, na.rm = TRUE)
#IQR(dat$Average.Daily.PM2.5, na.rm = TRUE)

#selecting the top 25 counties with the highest pm 2.5 values 
p24 <- dat %>% select(county, state, Average.Daily.PM2.5) %>% arrange(desc(Average.Daily.PM2.5))%>%
              filter(Average.Daily.PM2.5 > 11.1)

#code for the average daily pm 2.5 values in ascending order with top 25 counties 
p25 <- p24 %>% ggplot(aes(x = reorder(county, Average.Daily.PM2.5), y = Average.Daily.PM2.5, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Highest Average Daily PM 2.5") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Average Daily PM 2.5 micrograms per cubic meter") 

#adding state abbreviations to county line plot
p25 + scale_x_discrete("U.S. County", labels = c("Kern" = "Kern (CA)", "Kings" = "Kings (CA)", "San Bernardino" = "San Bernardino (CA)", "Tulare" = "Tulare (CA)", "Fresno" = "Fresno (CA)", "Riverside" = "Riverside (CA)", "San Joaquin" = "San Joaquin (CA)", "Plumas" = "Plumas (CA)", "Allegheny" = "Allegheny (PA)", "Stanislaus" = "Stanislaus (CA)", "Imperial" = "Imperial (CA)", "Pinal" = "Pinal (AZ)", "Jefferson" = "Jefferson (OH)", "Lebanon" = "Lebanon (PA)", "Madera" = "Madera (CA)", "DeKalb" = "DeKalb (GA)", "Los Angeles" = "Los Angeles (CA)", "Merced" = "Merced (CA)", "Lemhi" = "Lemhi (ID)", "Butler" = "Butler (OH)", "Shelby" = "Shelby (AL)", "Lancaster" = "Lancaster (PA)", "Wayne" = "Wayne (MI)", "Berks" = "Berks (PA)", "Montgomery" = "Montgomery (PA)"))


```

## Analysis 2.14: Median Household Income

```{r}
#histogram of median household income
p_income <- dat %>% ggplot(aes(x = Median.Household.Income)) +
              geom_histogram(binwidth = 1000, color = "black", fill = "gray") + 
              theme_bw()+ 
              ggtitle("Distribution of Median Household Income in the United States") +
              theme(plot.title = element_text(hjust = 0.5)) +
              xlab("Median Household Income") 

#addition a line where the median value is located 
p_income + geom_vline(aes(xintercept = median(Median.Household.Income, na.rm = TRUE)), col='dark red',size=1) 

#code for distribution statistics 
#median(dat$Median.Household.Income, na.rm = TRUE)
#IQR(dat$Median.Household.Income, na.rm = TRUE)


#code to determine the top 25 counties with the lowest median household income 
p26 <- dat %>% select(county, state, Median.Household.Income) %>% arrange(Median.Household.Income)%>%
              filter(Median.Household.Income < 31221)

#line plot in decsending order of median household income with the top 25 counties
p27 <- p26 %>% ggplot(aes(x = reorder(county, -Median.Household.Income), y = Median.Household.Income, group = 1)) +
            geom_point() +
            geom_line() + 
            theme_bw() + 
            ggtitle("Top 25 Counties With the Lowest Median Household Income") +
            theme(plot.title = element_text(hjust = 0.5)) + 
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            xlab("U.S County") + ylab("Median Household Income") 

#adding state abbreviations to county line plot
p27 + scale_x_discrete("U.S. County", labels = c("Clay" = "Clay (GA) & (KY)", "Holmes" = "Holmes (MS)", "Buffalo" = "Buffalo (SD)", "Taylor" = "Taylor (GA)", "Humphreys" = "Humphreys (MS)", "Greene" = "Greene (AL)", "Owsley" = "Owsley (KY)", "McDowell" = "McDowell (WV)", "Sumter" = "Sumter (AL)", "Quitman" = "Quitman (MS)", "Perry" = "Perry (AL)", "Leflore" = "Leflore (MS)", "Harlan" = "Harlan (KY)", "Clay" = "Clay (KY) & (GA)", "Coahoma" = "Coahoma (MS)", "Bolivar" = "Bolivar (MS)", "Lee" = "Lee (AR)", "Jefferson" = "Jefferson (MS)", "Lee" = "Lee (KY)", "Bell" = "Bell (KY)", "Issaquena" = "Issaquena (MS)", "Wilcox" = "Wilcox (MS)", "Washington" = "Washington (MS)", "Sharkey" = "Sharkey (MS)", "Johnson" = "Johnson (KY)"))

```

Outcome:
Our two primary outcomes were life expectancy and drug overdose mortality rate per 100,000 population. We created visualizations of the distribution of each outcome across the United States, by state, and then we determined the top 25 counties with the lowest life expectancy and the top 25 counties with the highest drug overdose mortality rate. We prioritized the histogram and line plots of the 25 counties for each predictor because our overall predictions were using county-level data. Depending on the distribution of the histogram of the variable of interest, the mean and the standard deviation were calculated, or the median and the interquartile range were calculated. We chose these metrics as these are the standard ways to summarize the distribution of variable. As we had county level data, we wanted to ensure that our analyses had a focus at the country level. there were no major changes to our initial ideas. We also wanted to show the relationship between our two main outcome variables, and that is why the scatter plot with the trend line was generated. 

# Analysis 3: Prediction models

## Analysis 3.1: linear regression

## The goal of this section was to identify the machine learning model that best predicts the outcomes of interest 

The built environment and place-based policies play important roles in life expectancy. They are also important factors in many of the epidemics facing America, including drug-related overdose mortality. It would be helpful to know what factors related to county-level characteristics help to predict differences in life expectancies and drug-related overdose mortality. 

Approach: To run a simple linear regression model to understand what covariates might be signficant and then to use correlation plots to pick up on potential collinearities and identify modifiable predictors that are still highly correlated with the outcomes of interest to include in the machine learning prediction models. 

Inspiration: Class sessions on logistic models, Naive Bayes, and Decision trees informed the prediction modelling approach. The outcomes were dichotomized as being below or above average life expectancy and drug overdose mortality respectively. 
```{r}

# prediction model

# We will be using data from the county health rankings to explore predictors for life expectancy using the following variables:

# Fair or Poor Health
# Average number of physically unhealthy days
# Average number of mentally unhealthy days
# low birthweight
# smoking status
# adults with obesity
# food environment index
# mental health provider rate 
# percent completed high school
# percent completed college
# percent unemployed
# violent crime rate 
# average daily PM 2.5
# Severe housing problems
# Median Household Income

# subset data to use for prediction model 

pred_dat <- dat %>% select(Life.Expectancy, `%.Fair.or.Poor.Health`, `Average.Number.of.Physically.Unhealthy.Days`,	`Average.Number.of.Mentally.Unhealthy.Days`,	`%.Low.birthweight`,	`%.Smokers`, `%.Adults.with.Obesity`,	`%.Food.Insecure`,	`%.Physically.Inactive`,	`%.Excessive.Drinking`,	`%.Uninsured`, `Primary.Care.Physicians.Rate`,	Mental.Health.Provider.Rate,	`%.Completed.High.School`, `%.Some.College`,	`%.Unemployed`,	Median.Household.Income, Violent.Crime.Rate, Average.Daily.PM2.5, `%.Severe.Housing.Problems`) %>% arrange(Life.Expectancy) %>% drop_na()

# create correlation plot 

library(GGally)
library(ggplot2)
corr <- cor(pred_dat)
ggcorr(corr, 
       label = T, 
       label_size = 2,
       label_round = 2,
       hjust = 1,
       size = 3, 
       color = "royalblue",
       layout.exp = 5,
       low = "red", 
       mid = "white", 
       high = "blue",
       name = "Correlation")

# We will be using data from the county health rankings to explore predictors for the drug-related overdose mortality outcomes using the following variables:

# Frequent.Physical.Distress
# Frequent.Mental.Distress
# percentage of Smokers
# percentage Food Insecure
# percentage Physically Inactive
# percentage Excessive Drinking
# percentage Uninsured
# Primary.Care.Physicians.Rate
# Mental.Health.Provider.Rate
# percentage Completed.High.School
# percentage completed Some.College
# percentage Unemployed
# SES, Median.Household.Income
# Violent.Crime.Rate
# percentage Insufficient.Sleep
# percentage Severe.Housing.Problems
# Life.Expectancy

# subset prediction data 

pred_dat2 <- dat %>% select(Drug.Overdose.Mortality.Rate,`%.Frequent.Physical.Distress`, `%.Frequent.Mental.Distress`,	`%.Smokers`, `%.Food.Insecure`,	`%.Physically.Inactive`,	`%.Excessive.Drinking`,	`%.Uninsured`, `Primary.Care.Physicians.Rate`,	Mental.Health.Provider.Rate,	`%.Completed.High.School`, `%.Some.College`,	`%.Unemployed`,	Median.Household.Income, Violent.Crime.Rate, `%.Insufficient.Sleep`, `%.Severe.Housing.Problems`, Life.Expectancy) %>% arrange(Drug.Overdose.Mortality.Rate) %>% drop_na()

# correlation plot
corr2 <- cor(pred_dat2)
ggcorr(corr2, 
       label = T, 
       label_size = 2,
       label_round = 2,
       hjust = 1,
       size = 3, 
       color = "royalblue",
       layout.exp = 5,
       low = "red", 
       mid = "white", 
       high = "blue",
       name = "Correlation")

# fit linear prediction model for life expectancy 

pred_fit <- glm(Life.Expectancy ~ ., data=pred_dat)
summary(pred_fit)

# fit linear prediction model for overdose mortality 

pred_fit2 <- glm(Drug.Overdose.Mortality.Rate ~ ., data=pred_dat2)
summary(pred_fit2)

# identify average life expectancy and overdose mortality to dichotomize variables 
avg.overdose <- mean(pred_dat2$Drug.Overdose.Mortality.Rate)
avg.overdose
avg.life <- mean(pred_dat$Life.Expectancy)
avg.life

# create data sets with dichotomized variables for outcomes of interest

pred_dat <- pred_dat %>% mutate(life_exp = ifelse(Life.Expectancy >= avg.life, 0, 1))

pred_dat2 <- pred_dat2 %>% mutate(overdose = ifelse(Drug.Overdose.Mortality.Rate >= avg.overdose, 1, 0))
```

## Analysis 3.2: machine learning models

```{r}
# prediction models 

library(tidyverse)
library(splitstackshape)
library(caret)
library(e1071)
library(pROC)
library(dplyr)
library(ggplot2)
remotes::update_packages("rlang")

set.seed(1)

# come up with training and test sets 
x <- stratified(pred_dat, "life_exp", 0.7, keep.rownames = TRUE)
train_set <- x %>% dplyr::select(-rn)
train_index <- as.numeric(x$rn)
test_set <- pred_dat[-train_index,]

dim(train_set)
dim(test_set)

## Logistic regression model (covariates selected based on relevance to policy/decision-making and correlation to outcomes)
fit <-  glm(life_exp ~ `%.Food.Insecure` +	`%.Uninsured` + `%.Some.College` +	`%.Unemployed` +	Median.Household.Income + Violent.Crime.Rate + Average.Daily.PM2.5 + `%.Severe.Housing.Problems`, data= train_set, family="binomial")

p_hat <- predict(fit, newdata = test_set, type="response")
y_hat <- ifelse(p_hat > 0.5, 1, 0)

# confusion matrix
CF <- confusionMatrix(data = as.factor(y_hat), reference = as.factor(test_set$life_exp), positive = '1')
CF

## Naive Bayes model
set.seed(1)
nb_fit <-  naiveBayes(life_exp ~ `%.Food.Insecure` +	`%.Uninsured` + `%.Some.College` +	`%.Unemployed` +	Median.Household.Income + Violent.Crime.Rate + Average.Daily.PM2.5 + `%.Severe.Housing.Problems`, data = train_set)

p_hat_nb <- predict(nb_fit, test_set, type = "raw")[,2]
y_hat_nb <- factor(ifelse(p_hat_nb > 0.5, 1, 0))

# confusion matrix
CFnb <- confusionMatrix(data = as.factor(y_hat_nb), reference = as.factor(test_set$life_exp), positive = '1')
CFnb

## decision tree model
library(rpart)
set.seed(1)
train_set <- train_set  %>% dplyr::select(., life_exp, `%.Food.Insecure` ,	`%.Uninsured` , `%.Some.College` ,	`%.Unemployed`,	Median.Household.Income , Violent.Crime.Rate , Average.Daily.PM2.5 , `%.Severe.Housing.Problems`)
tree_fit <- rpart(as.factor(life_exp) ~ ., data = train_set)
tree_preds <- predict(tree_fit, newdata = test_set, type = "class")

# confusion matrix
CFtree <- confusionMatrix(data = as.factor(tree_preds), reference = as.factor(test_set$life_exp), positive='1')
CFtree

# logistic life exp

roc_log <- roc(test_set$life_exp, y_hat)

# nb life exp

roc_nb <- roc(as.factor(test_set$life_exp), p_hat_nb)

# decision tree life exp
tree_preds <- predict(tree_fit, newdata = test_set, type = "prob")[,2]
roc_dt <- roc(test_set$life_exp, tree_preds)


# ROCs for life expectancy

ggroc(list("Logistic" = roc_log, "Naive Bayes" = roc_nb, "Decision Tree" = roc_dt)) +
  theme(legend.title = element_blank()) +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "black", linetype = "dashed") +
  xlab("Sensitivity") +
  ylab("Specificity")+
  ggtitle("Compare ROCs for Life Expectancy")+
  theme(plot.title = element_text(hjust = 0.5))

# overdose mortality

# creat training and test sets

x2 <- stratified(pred_dat2, "overdose", 0.7, keep.rownames = TRUE)
train_set2 <- x2 %>% dplyr::select(-rn)
train_index2 <- as.numeric(x2$rn)
test_set2 <- pred_dat2[-train_index2,]

dim(train_set2)
dim(test_set2)

## Logistic regression model (covariates selected based on relevance to policy/decision-making and correlation to outcomes)
fit2 <-  glm(overdose ~ `%.Food.Insecure`+	`%.Uninsured`+	`%.Some.College`+	`%.Unemployed`+	Median.Household.Income+ Violent.Crime.Rate+ `%.Insufficient.Sleep`, `%.Severe.Housing.Problems`, data= train_set2, family="binomial")

p_hat2 <- predict(fit2, newdata = test_set2, type="response")
y_hat2 <- ifelse(p_hat2 > 0.5, 1, 0)

# confusion matrix
CF2 <- confusionMatrix(data = as.factor(y_hat2), reference = as.factor(test_set2$overdose), positive = '1')
CF2

## Naive Bayes model
set.seed(1)
nb_fit2 <-  naiveBayes(overdose ~ `%.Food.Insecure`+	`%.Uninsured`+	`%.Some.College`+	`%.Unemployed`+	Median.Household.Income+ Violent.Crime.Rate+ `%.Insufficient.Sleep`+ `%.Severe.Housing.Problems`, data = train_set2)

p_hat_nb2 <- predict(nb_fit2, test_set2, type = "raw")[,2]
y_hat_nb2 <- factor(ifelse(p_hat_nb2 > 0.5, 1, 0))

# confusion matrix
CFnb2 <- confusionMatrix(data = as.factor(y_hat_nb2), reference = as.factor(test_set2$overdose), positive = '1')
CFnb2

## decision tree model
library(rpart)
set.seed(1)
train_set2 <- train_set2  %>% dplyr::select(., overdose, `%.Food.Insecure`,	`%.Uninsured`,	`%.Some.College`,	`%.Unemployed`,	Median.Household.Income, Violent.Crime.Rate, `%.Insufficient.Sleep`, `%.Severe.Housing.Problems`)
tree_fit2 <- rpart(as.factor(overdose) ~ ., data = train_set2)
tree_preds2 <- predict(tree_fit2, newdata = test_set2, type = "class")

# confusion matrix
CFtree2 <- confusionMatrix(data = as.factor(tree_preds2), reference = as.factor(test_set2$overdose), positive='1')
CFtree2

# logistic overdose

roc_log2 <- roc(test_set2$overdose, y_hat2)

# nb overdose

roc_nb2 <- roc(as.factor(test_set2$overdose), p_hat_nb2)

# decision tree overdose
tree_preds2 <- predict(tree_fit2, newdata = test_set2, type = "prob")[,2]
roc_dt2 <- roc(test_set2$overdose, tree_preds2)

# ROCs overdose

ggroc(list("Logistic" = roc_log2, "Naive Bayes" = roc_nb2, "Decision Tree" = roc_dt2)) +
  theme(legend.title = element_blank()) +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "black", linetype = "dashed") +
  xlab("Sensitivity") +
  ylab("Specificity")+
  ggtitle("Compare ROCs for Overdose Mortality")+
  theme(plot.title = element_text(hjust = 0.5))

```

## Analysis 3.3: identifying the best model and most important predictors

```{r}
# identifying best model 

# areas under the curve 
auc(roc_log)
auc(roc_log2)
auc(roc_nb)
auc(roc_nb2)
auc(roc_dt)
auc(roc_dt2)

# Variable importance from decision tree (life expectancy)

dat_imp <- data.frame(imp = tree_fit$variable.importance)
dat_imp <- dat_imp %>% 
  tibble::rownames_to_column() %>% 
  dplyr::rename("variable" = rowname) %>% 
  dplyr::arrange(imp) %>%
  dplyr::mutate(variable = forcats::fct_inorder(variable))
ggplot2::ggplot(dat_imp) +
  geom_col(aes(x = variable, y = imp),
           col = "black", show.legend = FALSE) +
  coord_flip() +
  scale_fill_grey() +
  theme_bw() + 
  ggtitle("Variable Importance: Life Expectancy (Decision Tree)")

# Variable importance from decision tree (overdose)

dat_imp2 <- data.frame(imp = tree_fit2$variable.importance)
dat_imp2 <- dat_imp2 %>% 
  tibble::rownames_to_column() %>% 
  dplyr::rename("variable" = rowname) %>% 
  dplyr::arrange(imp) %>%
  dplyr::mutate(variable = forcats::fct_inorder(variable))
ggplot2::ggplot(dat_imp2) +
  geom_col(aes(x = variable, y = imp),
           col = "black", show.legend = FALSE) +
  coord_flip() +
  scale_fill_grey() +
  theme_bw() + 
  ggtitle("Variable Importance: Overdose Mortality (Decision Tree)")

```

Table for life Expectancy Prediction models

| Model |Accuracy |Sensitivity | Specificity | AUC |
| :---                |    :----:                     |  :----:                  | :----:           | :----:         |       ---:            |
|Logistic| `r round(CF$overall["Accuracy"], 3)`| `r round(CF$byClass["Sensitivity"], 3)`|`r round(CF$byClass["Specificity"], 3)`| `r round(auc(roc_log), 3)`|
|Naive Bayes| ``r round(CFnb$overall["Accuracy"], 3)`| `r round(CFnb$byClass["Sensitivity"], 3)`|`r round(CFnb$byClass["Specificity"], 3)`| `r round(auc(roc_nb), 3)`|
|Decision Tree | `r round(CFtree$overall["Accuracy"], 3)`| `r round(CFtree$byClass["Sensitivity"], 3)`|`r round(CFtree$byClass["Specificity"], 3)`| `r round(auc(roc_dt), 3)`|

Table for Overdose Mortality Prediction models

| Model |Accuracy |Sensitivity | Specificity | AUC |
| :---                |    :----:                     |  :----:                  | :----:           | :----:         |       ---:            |
|Logistic| `r round(CF2$overall["Accuracy"], 3)`| `r round(CF2$byClass["Sensitivity"], 3)`|`r round(CF2$byClass["Specificity"], 3)`| `r round(auc(roc_log2), 3)`|
|Naive Bayes| ``r round(CFnb2$overall["Accuracy"], 3)`| `r round(CFnb2$byClass["Sensitivity"], 3)`|`r round(CFnb2$byClass["Specificity"], 3)`| `r round(auc(roc_nb2), 3)`|
|Decision Tree | `r round(CFtree2$overall["Accuracy"], 3)`| `r round(CFtree2$byClass["Sensitivity"], 3)`|`r round(CFtree2$byClass["Specificity"], 3)`| `r round(auc(roc_dt2), 3)`|

# Analysis 4: Mapping

```{r, warning=FALSE}
# Loading necessary libraries
library(openxlsx)
library(tidyverse)
library(biscale)
library(rgdal)
library(ggplot2)
library(sf)
library(sp)
library(tibble)
library(cowplot)
library(ggsn)
library(rgeos)
library(raster)
library(staplr)
library(grid)

# Read in county shapefile from US Census TIGER/LINE
county_shp <- rgdal::readOGR(".", layer = "cb_2018_us_county_20m",
                            verbose = F)

# Convert county shapefile to sf object
county_sf <- as(county_shp, "sf")

# Read in state shapefile from US Census TIGER/LINE
state_shp <- rgdal::readOGR(".", layer = "cb_2018_us_state_20m",
                            verbose = F)

# Convert state shapefile to sf object
state_sf <- as(state_shp, "sf")

# Remove Puerto Rico from the state shapefile
state_sf <- state_sf %>% filter(STATEFP != "72")

# Create a new variable that merges the state and county FIPS codes to a single identifier
county_sf <- county_sf %>% mutate(fips = paste(STATEFP, COUNTYFP, sep = ""))

# Reduce the size of the county health rankings data to only the relevant columns
dat_subset <- dplyr::select(dat2, c(fips, Drug.Overdose.Mortality.Rate, Life.Expectancy))

# Join the county health rankings data to the county shapefile
county_sf <- left_join(county_sf, dat_subset, by = c("fips" = "fips"))

# Remove all US territories and non-relevant FIPS codes from the county shapefile
county_sf <- county_sf %>% filter(STATEFP != "14" & STATEFP != "03" & STATEFP != "43" & STATEFP != "52" & STATEFP != "78" & STATEFP != "79" & STATEFP != "74" & STATEFP != "72" & STATEFP != "69" & STATEFP != "70" & STATEFP != "95" & STATEFP != "71" & STATEFP != "76" & STATEFP != "68" & STATEFP != "89" & STATEFP != "67" & STATEFP != "86" & STATEFP != "84" & STATEFP != "66" & STATEFP != "64" & STATEFP != "81" & STATEFP != "60" & COUNTYFP != "016")

# Drop counties in the shapefile that have missing values for drug overdose mortality rate
county_sf_no_od_na <- county_sf %>% filter(is.na(Drug.Overdose.Mortality.Rate) == FALSE)
```

```{r}
# Create the 3-quantile cutoffs based on the distributions of overdose mortality rate and life expectancy
od_le_class <- bi_class(county_sf_no_od_na, x = Drug.Overdose.Mortality.Rate, y = Life.Expectancy, style = "quantile", dim = 3)

# Bivariate map of drug overdose mortality rate and life expectancy excluding Alaska and Hawaii, which will be inset later
map_od_le <- ggplot() +
  geom_sf(data = filter(od_le_class, STATEFP != "15" & STATEFP != "02"), mapping = aes(fill = bi_class),
          color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkViolet", dim = 3) +
  bi_theme() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        rect = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA)) + 
  geom_sf(data = filter(county_sf, STATEFP != "15" & STATEFP != "02"), fill = NA, color = "gray50", size = 0.1, show.legend = FALSE) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA)) + 
  geom_sf(data = filter(state_sf, STATEFP != "15" & STATEFP != "02"), fill = NA, color = "black",   size = 0.15, show.legend = FALSE) + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))

# Bivariate map of drug overdose mortality rate and life expectancy for Hawaii inset
map_od_le_hi <- ggplot() +
  geom_sf(data = filter(od_le_class, STATEFP == "15"), mapping = aes(fill = bi_class),
          color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkViolet", dim = 3) +
  bi_theme() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        rect = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA)) + 
  geom_sf(data = filter(county_sf, STATEFP == "15"), fill = NA, color = "gray50", size = 0.1, show.legend = FALSE) + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA)) + 
  geom_sf(data = filter(state_sf, STATEFP == "15"), fill = NA, color = "black", size = 0.15, show.legend = FALSE) + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))

# Bivariate map of drug overdose mortality rate and life expectancy for Alaska inset
map_od_le_ak <- ggplot() +
  geom_sf(data = filter(od_le_class, STATEFP == "02"), mapping = aes(fill = bi_class),
          color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkViolet", dim = 3) +
  bi_theme() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        rect = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA)) + 
  geom_sf(data = filter(state_sf, STATEFP == "02"), fill = NA, color = "black", size = 0.15, show.legend = FALSE) + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA)) + 
  geom_sf(data = filter(county_sf, STATEFP == "02"), fill = NA, color = "gray50", size = 0.1, show.legend = FALSE) + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))

legend_od_le <- bi_legend(pal = "DkViolet",
                    dim = 3,
                    xlab = "Higher OD Rate",
                    ylab = "Higher LE",
                    size = 8)

# Text box indicating the value cutoffs for the quantiles for each variable
annotation <- ggplot() + 
  annotation_custom(grobTree(textGrob("Quantile Cutoffs\nOD Mortality (deaths/100,000): 3.82, 15.84, 24.92, 126.73\nLE (years): 64.41, 76.19, 78.70, 113.46", x = 0.38,  y = 0.05, hjust = 0, gp = gpar(col = "black", fontsize = 8)))) + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))

# Use cowplot to assemble the map components
od_le_bivariate <- ggdraw() +
  draw_plot(map_od_le, 0, 0, 1, 1) +
  draw_plot(legend_od_le, 0.75, 0.03, 0.25, 0.25) +
  draw_plot(map_od_le_hi, 0.2, 0.01, 0.2, 0.2) + 
  draw_plot(map_od_le_ak, -0.07, -0.6, 1.6, 1.6) + 
  draw_plot(annotation)

od_le_bivariate


```

The map shows the joint spatial distribution of drug overdose mortality rate and life expectancy by county. Counties in white have missing information for drug overdose mortality rate. Quantile cutoffs are the most appropriate choice for spatial classification of epidemiological rate data fro presentation, especially when variables such as drug overdose mortality rate are skewed (Brewer and Pickle, 2002).

Areas of high life expectancy and high drug overdose mortality rate are concentrated in the northeast corridor and south Florida. Areas of low life expectancy and low overdose mortality rates are concentrated southern Appalachia and the rural West. Areas of high life expectancy and low overdose mortality rates are concentrated in upper Midwest, east Texas, and the west coast, including Hawaii.

Brewer CA, Pickle L. Evaluation of methods for classifying epidemiological data on choropleth maps in series. Ann Assoc Am Geogr. 2002;92(4):662681.



